### 2020-05-22 serial.tex
tweak

The serial interval is defined as the time between when an infector becomes symptomatic and when an infectee becomes symptomatic \citep{svensson2007note}.
Previous studies have often expressed serial intervals $\tau$ in the form (\fref{diagram}A):
\begin{equation}
\tau = (\gtime + x_1) - x_0
\end{equation}
where $x_0$ and $x_1$ represent incubation periods of an infector and an infectee, respectively, and $\gtime$ represents the generation interval.
These studies assumed that $x_0$ and $x_1$ follow the same distributions and concluded that the serial and generation intervals have the same mean \citep{svensson2007note,klinkenberg2011correlation,champredon2018equivalence, britton2019estimation};
however, this conclusion is based on the implicit assumption that incubation times, $x_0$ and $x_1$, as well as generation interval, $\gtime$, are intrinsic to individuals (and not dependent on disease dynamics) --- something that is true of forward but not backward incubation times.
We refer to this definition as the intrinsic serial interval (\fref{diagram}A).

\begin{figure}[!th]
\includegraphics[width=\textwidth]{serial_guide.pdf}
\caption{
\textbf{Illustration of intrinsic, forward and backward serial intervals.}
(A) The intrinsic serial interval for a cohort of individuals infected at time $\psymp$.
In this case, $x_0$ follows the intrinsic incubation period distribution;
$\gtime$ follows the intrinsic generation-interval distribution;
and $x_1$ follows the intrinsic incubation period distribution.
(B) The forward serial interval for a cohort of infectors who became symptomatic at time $\psymp$.
In this case, $x_0$ follows the backward incubation period distribution;
$\gtime$ follows the forward generation-interval distribution;
and $x_1$ follows the forward incubation period distribution.
(C) The backward serial interval for a cohort of infectees who became symptomatic at time $\ssymp$.
In this case, $x_0$ follows the forward incubation period distribution;
$\gtime$ follows the backward generation-interval distribution;
and $x_1$ follows the backward incubation period distribution.
}
\label{fig:diagram}
\end{figure}

To correctly link realized serial intervals to the renewal process between cases based on symptom onset dates, we must use the forward serial interval (i.e., use the perspective of a cohort of infectors that share the same symptom onset time). 
Given that an infector became symptomatic at time $\psymp$, to calculate the forward serial interval we first go backward in time by asking when the infector was infected, and then go forward in time by asking first when the infectee was infected and then when they became symptomatic.
In \fref{diagram}B, we see that $x_0$ follows the backward incubation period distribution of the cohort of infectors who became symptomatic at time $\psymp$;
$\gtime$ follows the forward generation-interval distribution of the cohort of infectors who came infected at time $\psymp - x_0$;
and $x_1$ follows the forward incubation period distribution of the cohort of infectees who became infected at time $\psymp - x_0 + \gtime$.
Likewise, we can define the backward serial interval distribution for a cohort of infectees who became symptomatic at time $\ssymp$ (\fref{diagram}C).
This conceptual framework clearly demonstrates that the distributions of $x_0$, $\gtime$, and $x_1$ (and therefore the distributions of realized serial intervals) depend on the reference cohort and the corresponding cohort time.

@@

The total number of serial intervals can now be obtained by integrating over all possible infection times for the infector and the infectee:
\begin{equation}
\total (\psymp,\ssymp) = \int_{-\infty}^{\psymp} \int_{\pinf}^{\ssymp} \Rc (\pinf) i(\pinf) h_{\pinf}(\psymp-\pinf, \sinf - \pinf) \idist_{\sinf}(\ssymp - \sinf) \, \mathrm{d}\sinf\,\mathrm{d}\pinf.
\end{equation}
Then, the forward serial-interval distribution $f_\psymp(\tau)$ is given by the density of intervals of length $\tau$ starting at \psymp, relative to the total number of serial intervals starting at \psymp: 
\begin{equation}
f_\psymp(\tau) = 
\frac{\total(\psymp, \psymp+\tau)}{\int \total(\psymp, \psymp+\tau') d\tau'}.
\label{eq:fserial}
\end{equation}
Likewise, the backward serial-interval distribution $b_\ssymp(\tau)$ is given by the density of intervals of length $\tau$ ending at \ssymp, relative to the total number of serial intervals ending at \ssymp: 
\begin{equation}
b_\ssymp(\tau) = 
\frac{\total(\ssymp-\tau, \ssymp)}{\int \total(\ssymp-\tau', \ssymp) d\tau'}.
\label{eq:bserial}
\end{equation}
This framework allows us to understand changes in the realized serial intervals for any deterministic epidemic models.

@@

\begin{table}[!th]
\begin{center}
\begin{tabular}{|l|l|r|}
\hline
Parameter & Values & Source\\
\hline
Mean intrinsic incubation period & 5.5 days & \cite{lauer2020incubation} \\
SD intrinsic incubation period & 2.4 days & \cite{lauer2020incubation} \\
Mean intrinsic generation interval & 5 days & \cite{ferretti2020quantifying} \\
SD intrinsic generation interval & 2 days & \cite{ferretti2020quantifying} \\
\hline
\end{tabular}
\end{center}
\caption{
\textbf{Parameter values used for simulations.}
The intrinsic generation-interval distribution is parameterized using a log-normal distribution with log mean $\mu_G=1.54$ and log standard deviation $\sigma_G=0.37$.
The forward incubation period distribution is parameterized using a log-normal distribution with log mean $\mu_I=1.62$ and log standard deviation $\sigma_I=0.42$.
The joint probability distribution is modeled using a multivariate log-normal distribution with correlations (on the log scale) $\rho=-0.5, 0, 0.5$.
}
\end{table}

@@

We simulate the renewal equation model using a discrete-time approximation:
\begin{equation}
\begin{aligned}
i(t) &= \Ro S(t-\Delta t) \sum_{m=1}^{\ell} i(t-m \Delta t) \hat{\gdist}(m \Delta t) \\
S(t) &= S(t-\Delta t) - i(t)
\end{aligned}
\end{equation}
where $\hat{\gdist}$ is a discrete-time intrinsic generation-interval distribution that satisfies the following:
\begin{equation}
\hat{\gdist}(m \Delta t) = \frac{\gdist(m \Delta t)}{\sum_{i=1}^\ell \gdist(m \Delta t)}, \quad m=1, \dots, \ell
\end{equation}
The continuous-time intrinsic generation-interval distribution is parameterized using a log-normal distribution (Table 1). We define the intrinsic incubation period distribution in a similar manner:
\begin{equation}
\hat{\idist}(m \Delta t) = \frac{\idist(m \Delta t)}{\sum_{i=1}^\ell \idist(m \Delta t)}, \quad m=1, \dots, \ell,
\end{equation}
where its continuous-time analog is also based on a log-normal distribution.
For simplicity, we assume that the forward incubation periods and intrinsic generation intervals are independent:
\begin{equation}
\hat{h}(m \Delta t, \Delta t n) = \hat{\idist}(m \Delta t)\hat{\gdist}(\Delta t n), \quad m,n=1, \dots, \ell.
\end{equation}
We use $\Delta t = 0.025\,\textrm{days}$ and $\ell=2001$ for discretization steps.

We initialize the simulation with population size $N$=40,000 as follows:
\begin{equation}
\begin{aligned}
i(m \Delta t) &= C \exp(r m \Delta t), \quad m=1, \dots, \ell\\
S(m \Delta t) &= N - \sum_{n=1}^m i(m \Delta t), \quad m=1, \dots, \ell\\
\end{aligned}
\end{equation}
where $C$ is chosen such that $\sum_{n=1}^\ell i(m \Delta t)=10$.
These initial conditions allow the model to follow exponentially growth from time $\Delta t (\ell + 1)$ without any transient behaviors.
Results presented in the main text show simulations beginning from time $\Delta t (\ell + 1)$.

@@

We use a simulation-based approach to compare the estimates of \Ro based on the serial- and generation-interval distributions. 
To do so, we model the intrinsic generation-interval distribution and the incubation period using a multivariate log-normal distribution with log means $\mu_G, \mu_I$, log standard variances $\sigma_G^2, \sigma_I^2$, and log-scale correlation $\rho$;
the multivariate log-normal distribution is parameterized based on parameter estimates for COVID-19 (Table 1).
We construct forward serial intervals during the exponential growth period as follows:
\begin{equation}
F_i = -X_{0,i} + (G_i|X_{0,i}) + X_{1,i},
\end{equation}
where the backward incubation period $X_{0,i}$ of an infector is simulated by drawing random log-normal samples $Y_i$ with log mean $\mu_I$ and log variance $\sigma_I^2$ and resampling $Y_i$, each weighted by the inverse of the exponential growth function $\exp(-rY_i)$;
the intrinsic generation interval conditional on the incubation period of the infector $(G_i|X_{0,i})$ is drawn from a log-normal distribution with log mean $\mu_G + \sigma_G \rho (\log(X_{0,i}) - \mu_I)/\sigma_I$ and log variance $\sigma_G^2 (1-\rho^2)$;
the forward incubation period $X_{1,i}$ of an infectee is drawn from a log-normal distribution with log mean $\mu_I$ and log variance $\sigma_I^2$.
We then calculate the basic reproduction number \Ro using the empirical estimator:
\begin{equation}
\Ro = \frac{1}{\frac{1}{N}\sum_{i=1}^N \exp(- r F_i)}.
\end{equation}
We compare this with an estimate of \Ro based on the intrinsic serial-interval distribution in which the backward and the forward incubation periods are identically distributed \citep{svensson2007note,klinkenberg2011correlation,champredon2018equivalence, britton2019estimation}:
\begin{equation}
  \Rintrinsic = \frac{1}{\frac{1}{N}\sum_{i=1}^N \exp(- r Q_i)},
\end{equation}
where
\begin{equation}
Q_i = -Y_i + (G_i|Y_i) + X_{1,i}.
\end{equation}

>>>>

The serial interval is defined as the time between when an infector becomes symptomatic and when an infectee becomes symptomatic \citep{svensson2007note}.
Previous studies have often expressed serial intervals $\tau$ in the form (\fref{diagram}A):
\begin{equation}
\tau = (\gtime + x_2) - x_1
\end{equation}
where $x_1$ and $x_2$ represent incubation periods of an infector and an infectee, respectively, and $\gtime$ represents the generation interval.
These studies assumed that $x_1$ and $x_2$ follow the same distributions and concluded that the serial and generation intervals have the same mean \citep{svensson2007note,klinkenberg2011correlation,champredon2018equivalence, britton2019estimation};
however, this conclusion is based on the implicit assumption that incubation times, $x_1$ and $x_2$, as well as generation interval, $\gtime$, are intrinsic to individuals (and not dependent on disease dynamics) --- something that is true of forward but not backward incubation times.
We refer to this definition as the intrinsic serial interval (\fref{diagram}A).

\begin{figure}[!th]
\includegraphics[width=\textwidth]{serial_guide.pdf}
\caption{
\textbf{Illustration of intrinsic, forward and backward serial intervals.}
(A) The intrinsic serial interval for a cohort of individuals infected at time $\psymp$.
In this case, $x_1$ follows the intrinsic incubation period distribution;
$\gtime$ follows the intrinsic generation-interval distribution;
and $x_2$ follows the intrinsic incubation period distribution.
(B) The forward serial interval for a cohort of infectors who became symptomatic at time $\psymp$.
In this case, $x_1$ follows the backward incubation period distribution;
$\gtime$ follows the forward generation-interval distribution;
and $x_2$ follows the forward incubation period distribution.
(C) The backward serial interval for a cohort of infectees who became symptomatic at time $\ssymp$.
In this case, $x_1$ follows the forward incubation period distribution;
$\gtime$ follows the backward generation-interval distribution;
and $x_2$ follows the backward incubation period distribution.
}
\label{fig:diagram}
\end{figure}

To correctly link realized serial intervals to the renewal process between cases based on symptom onset dates, we must use the forward serial interval (i.e., use the perspective of a cohort of infectors that share the same symptom onset time). 
Given that an infector became symptomatic at time $\psymp$, to calculate the forward serial interval we first go backward in time by asking when the infector was infected, and then go forward in time by asking first when the infectee was infected and then when they became symptomatic.
In \fref{diagram}B, we see that $x_1$ follows the backward incubation period distribution of the cohort of infectors who became symptomatic at time $\psymp$;
$\gtime$ follows the forward generation-interval distribution of the cohort of infectors who came infected at time $\psymp - x_1$;
and $x_2$ follows the forward incubation period distribution of the cohort of infectees who became infected at time $\psymp - x_1 + \gtime$.
Likewise, we can define the backward serial interval distribution for a cohort of infectees who became symptomatic at time $\ssymp$ (\fref{diagram}C).
This conceptual framework clearly demonstrates that the distributions of $x_1$, $\gtime$, and $x_2$ (and therefore the distributions of realized serial intervals) depend on the reference cohort and the corresponding cohort time.

@@

The total number of serial intervals can now be obtained by integrating over all possible infection times for the infector and the infectee:
\begin{equation}
\total (\psymp,\ssymp) = \int_{-\infty}^{\psymp} \int_{\pinf}^{\ssymp} \Rc (\pinf) i(\pinf) h_{\pinf}(\psymp-\pinf, \sinf - \pinf) \idist_{\sinf}(\ssymp - \sinf) \, \mathrm{d}\sinf\,\mathrm{d}\pinf.
\end{equation}
Then, the forward serial-interval distribution $f_\psymp(\tau)$ is given by the density of intervals of length $\tau$ starting at \psymp, relative to the total number of serial intervals starting at \psymp: 
\begin{equation}
f_\psymp(\tau) = 
\frac{\total(\psymp, \psymp+\tau)}{\int \total(\psymp, \psymp+\tau') \dtau'}.
\label{eq:fserial}
\end{equation}
Likewise, the backward serial-interval distribution $b_\ssymp(\tau)$ is given by the density of intervals of length $\tau$ ending at \ssymp, relative to the total number of serial intervals ending at \ssymp: 
\begin{equation}
b_\ssymp(\tau) = 
\frac{\total(\ssymp-\tau, \ssymp)}{\int \total(\ssymp-\tau', \ssymp) \dtau'}.
\label{eq:bserial}
\end{equation}
This framework allows us to understand changes in the realized serial intervals for any deterministic epidemic models.

@@

\begin{table}[!th]
\begin{center}
\begin{tabular}{|l|l|r|}
\hline
Parameter & Values & Source\\
\hline
Mean intrinsic incubation period & 5.5 days & \cite{lauer2020incubation} \\
SD intrinsic incubation period & 2.4 days & \cite{lauer2020incubation} \\
Mean intrinsic generation interval & 5 days & \cite{ferretti2020quantifying} \\
SD intrinsic generation interval & 2 days & \cite{ferretti2020quantifying} \\
\hline
\end{tabular}
\end{center}
\caption{
\textbf{Parameter values used for simulations.}
The intrinsic generation-interval distribution is parameterized using a log-normal distribution with log mean $\mu_G=1.54$ and log standard deviation $\sigma_G=0.37$.
The intrinsic incubation period distribution is parameterized using a log-normal distribution with log mean $\mu_I=1.62$ and log standard deviation $\sigma_I=0.42$.
The joint probability distribution is modeled using a multivariate log-normal distribution with correlations (on the log scale) $\rho=-0.5, 0, 0.5$.
}
\end{table}

@@

We simulate the renewal equation model using a discrete-time approximation:
\begin{equation}
\begin{aligned}
i(t) &= \Ro S(t-\Delta t) \sum_{m=1}^{\tsub{m}{max}} i(t-m \Delta t) \hat{\gdist}(m \Delta t) \\
S(t) &= S(t-\Delta t) - i(t)
\end{aligned}
\end{equation}
where $\hat{\gdist}$ is a discrete-time intrinsic generation-interval distribution that satisfies the following:
\begin{equation}
\hat{\gdist}(m \Delta t) = \frac{\gdist(m \Delta t)}{\sum_{i=1}^\ell \gdist(m \Delta t)}, \quad m=1, \dots, \tsub{m}{max}.
\end{equation}
The continuous-time intrinsic generation-interval distribution is parameterized using a log-normal distribution (Table 1). We define the intrinsic incubation period distribution in a similar manner:
\begin{equation}
\hat{\idist}(m \Delta t) = \frac{\idist(m \Delta t)}{\sum_{i=1}^\ell \idist(m \Delta t)}, \quad m=1, \dots, \tsub{m}{max},
\end{equation}
where its continuous-time analog is also based on a log-normal distribution.
For simplicity, we assume that the forward incubation periods and intrinsic generation intervals are independent:
\begin{equation}
\hat{h}(m \Delta t, n \Delta t) = \hat{\idist}(m \Delta t)\hat{\gdist}(n \Delta t), \quad m,n=1, \dots, \tsub{m}{max}.
\end{equation}
We use $\Delta t = 0.025\,\textrm{days}$ and $\tsub{m}{max}=2001$ for discretization steps.

We initialize the simulation with population size $N$=40,000 as follows:
\begin{equation}
\begin{aligned}
i(m \Delta t) &= C \exp(r m \Delta t), \quad m=1, \dots, \tsub{m}{max}\\
S(m \Delta t) &= N - \sum_{n=1}^m i(n \Delta t), \quad m=1, \dots, \tsub{m}{max}\\
\end{aligned}
\end{equation}
where $C$ is chosen such that $\sum_{n=1}^{\tsub{m}{max}} i(m \Delta t)=10$.
These initial conditions allow the model to follow exponentially growth from time $\Delta t (\tsub{m}{max} + 1)$ without any transient behaviors.
Results presented in the main text show simulations beginning from time $\Delta t (\tsub{m}{max} + 1)$.

@@

We use a simulation-based approach to compare the estimates of \Ro based on the serial- and generation-interval distributions. 
To do so, we model the intrinsic generation-interval distribution and the incubation period using a multivariate log-normal distribution with log means $\mu_G, \mu_I$, log standard variances $\sigma_G^2, \sigma_I^2$, and log-scale correlation $\rho$;
the multivariate log-normal distribution is parameterized based on parameter estimates for COVID-19 (Table 1).
We construct forward serial intervals during the exponential growth period as follows:
\begin{equation}
F_i = -X_{1,i} + (G_i|X_{1,i}) + X_{2,i},
\end{equation}
where the backward incubation period $X_{1,i}$ of an infector is simulated by drawing random log-normal samples $Y_i$ with log mean $\mu_I$ and log variance $\sigma_I^2$ and resampling $Y_i$, each weighted by the inverse of the exponential growth function $\exp(-rY_i)$;
the intrinsic generation interval conditional on the incubation period of the infector $(G_i|X_{1,i})$ is drawn from a log-normal distribution with log mean $\mu_G + \sigma_G \rho (\log(X_{1,i}) - \mu_I)/\sigma_I$ and log variance $\sigma_G^2 (1-\rho^2)$;
the forward incubation period $X_{2,i}$ of an infectee is drawn from a log-normal distribution with log mean $\mu_I$ and log variance $\sigma_I^2$.
We then calculate the basic reproduction number \Ro using the empirical estimator:
\begin{equation}
\Ro = \frac{1}{\frac{1}{N}\sum_{i=1}^N \exp(- r F_i)}.
\end{equation}
We compare this with an estimate of \Ro based on the intrinsic serial-interval distribution in which the backward and the forward incubation periods are identically distributed \citep{svensson2007note,klinkenberg2011correlation,champredon2018equivalence, britton2019estimation}:
\begin{equation}
  \Rintrinsic = \frac{1}{\frac{1}{N}\sum_{i=1}^N \exp(- r Q_i)},
\end{equation}
where
\begin{equation}
Q_i = -Y_i + (G_i|Y_i) + X_{2,i}.
\end{equation}
